GEOG0149 - Week 8 Practical

To what extent is there residential segregation in English cities?

Background
Governments in Britain and other European countries have long espoused a commitment to creating ‘mixed communities’ in residential neighbourhoods. Most of the focus has been on two dimensions of residential mixing: (i) between socio-economic groups (often defined using social class or income variables) and (ii) between groups defined in terms of ethnicity, race, country of birth or ethno-religious background (a huge range of measures have been used to do this which makes comparison rather tricky).

In 2016, the Casey Review into opportunity and integration identified problems with divided communities as well as social, economic and cultural isolation in some parts of the UK. It argued that community division and a lack of mixing were having a wide range of adverse impacts on people and places. Much of the review’s focus was on relations between people from different ethnic and religious groups, although it also discussed class and economic divisions in some detail.

Partly in response to the review, in 2018 the UK Government published an Integrated Communities Strategy Green Paper. This was followed in early 2019 by an Integrated Communities Action Plan which has recently been evaluated.  The Green Paper and Plan draw explicitly on contact theories to build a case that mixed communities are more prosperous, healthy and pleasant places to live than those where there is less residential integration.  Again, most of the focus was on ethnicity, migration and the ethno-religious geography of England, with less discussion of how to foster integration or mixing between people from different socio-economic groups. Following the Strategy’s initial publication, five Local Authority Districts (LADs) — Blackburn with Darwen, Bradford, Peterborough, Walsall and Waltham Forest (click here for a map)— were selected as test Integration Areas. These LADs drew up local plans on improving integration in their area with help from central government.

Your task today is to explore some of the assumptions that lie behind the Integration Strategy. In particular, you are tasked with examining whether the five Integration Areas exhibit a high degree of residential segregation along the lines of ethnicity. We’ll also explore the geography of where groups live using relatively recent ethnicity data from the 2021 census.

Topics covered:
1. Computing and comparing segregation indices
2. Mapping ethnicity
3. Measuring clustering using local statistics

```{r}
# load the here library or managing file paths in projects
library(here)
# load 'MLID' package for calculating measures of segregation and inequality
library(MLID)
# load tidyverse package for data manipulation and visualiation
library(tidyverse)
# load sf package for handling spatial data
library(sf)
# load tmap for creating thematic maps
library(tmap)
# load spdep package for spatial dependence analysis 
library(spdep)
# load stargazer to created well-formatted output tables
library(stargazer)
```

```{r}
# read csv data containing ethnicity information or 4,301 Output Areas (OAs) in five local authority districts
ethnicity_OAs <- read_csv(here("Practical and Evidence Brief resources-20250311", "Ethnicity_OA_2021.csv"))
```

```{r}
# filter the ethnicity_OAs data to include only rows for Bradford
bradford_OAs <- filter(ethnicity_OAs, Local_Authority == "Bradford")
```

```{r}
# create a summary of Bradford ethnic composition data as percentages of the total OA population
bradford_summary <- bradford_OAs %>% 
  # for each ethnic group column from 'White_British' all the way to 'Other_groups',
  # calculate the percentage of each ethnic group's population relative to the total OA population
  summarise(
    across(White_British:Other_groups, ~ .x/Total_OA_pop*100)
  ) %>% 
  # add a new column call 'rowid' to uniquely identify each row
  mutate(rowid = row_number())
```

```{r}
# transform Bradford ethnicity data from wide to long format for easier plotting and analysis
bradford_summary_long <- pivot_longer(bradford_summary, cols = White_British:Other_groups, names_to = "Ethnicity", values_to = "Percent")
```

```{r}
# create a boxplot showing the distribution of ethnicity percentages across Bradford OAs
ggplot(bradford_summary_long, aes(Ethnicity, Percent)) +
  geom_boxplot() + # add boxplots for each ethnic group
  theme_bw() + # apply a clean black-and-white them
  ggtitle("Distribution of Ethnicity Percentages in Bradford OAs") + # plot title
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# calculate summary statistics (mean, median, maximum, minimum) for each ethnic group
bradford_summary_statistics <- bradford_summary_long %>% 
  group_by(Ethnicity) %>%  # group data y ethnicity
  summarise(
    mean = mean(Percent), # compute average percentage
    median = median(Percent), # compute median percentage
    maximum = max(Percent), # compute maximum percentage
    minimum = min(Percent) # compute minimum percentage
    ) %>%
  mutate(across(c(mean, median, maximum, minimum), # round all statistics to 1 decimal place
                ~ round(.x, 1)))
# inspect
# the median percentage of White British residents in Bradford's OAs is 77.5%
# the maximum percentage of Pakistani residents in Bradford's OAs is 93.4%
# the median percentage of Indian residents in Bradford's OAs is 1.1%
# the mean percentage of Black Caribbean residents in Bradford's OA is 0.5%
bradford_summary_statistics
```

```{r}
# load the Output Area (OA) chapefile containing geographic bounadaries
output_area_shapes <- st_read(
  here("Practical and Evidence Brief resources-20250311","OA_shapes_2021.shp")
)

# quick plot a simple map of the OA boundaries using tmap
qtm(output_area_shapes)
```

```{r}
# merge shapefile with Bradford's ethnicity data to create a spatial data frame using the OA21CD and OA_code column
# default merging will drop all row which do not appear in both the datasets that are being merged - only the Bradford OAs will be left after merging
bradford_map <- merge(output_area_shapes, bradford_OAs, by.x = "OA21CD", by.y = "OA_code")
```

```{r}
# recalculate percentage-based ethnicity columns to the Bradford spatial data frame
bradford_map <- mutate(bradford_map, across(White_British:Other_groups, ~.x/Total_OA_pop*100,.names = paste("p", "{.col}", sep = "_")))
```

```{r}
# create thematic map visualising percentages of White British and Pakistani groups in Bradford Output Areas
tm_shape(bradford_map) +
  tm_fill(col = c("p_White_British", "p_Pakistani"), # columns to visualise as percentages
          title = c("% White British", "% Pakistani"),
          style = "fixed", # use fixed intervals for colour breaks
          breaks = c(0, 10, 20, 50, 100)) + # defined intervals (0-10, 10-20, 20-50, 50-100)
  tm_borders(col = "darkgrey",
             alpha = .25) +
  tm_scale_bar(position = c("LEFT", "BOTTOM")) +
  tm_compass(position = c("RIGHT", "TOP")) +
  tm_layout(main.title = "% of White British and Pakistani Populations in Bradford OAs",
            main.title.position = c("center", "top"),
            main.title.size = 1.2,
            main.title.fontface = "bold",
            legend.frame = TRUE,
            legend.title.size = 1)
```

```{r}
# calculate the index of dissimilarity between Pakistani and White British groups in Bradford OAs
dissimilarity_index <- MLID::id(bradford_OAs,
                          vars = c("Pakistani", "White_British"),
                          )

# inspect
# the dissimilarity value of 0.76 is quite high and tells you the distribution of Pakistani and White British residents in Bradford is quite uneven, as we saw earlier in the maps
dissimilarity_index
```

```{r}
# create a new data frame combining ethnicity groups into broader categories
bradford_OAs_sum <- bradford_OAs %>% 
  mutate(Whites = White_British + Other_White) %>% # whites
  mutate(All_minorities = Indian + Pakistani + Bangladeshi + Chinese + African + Caribbean + Mixed + Other_groups) # all non-whites

dissimilarity_index_2 <- MLID::id(bradford_OAs_sum, vars = c("Whites", "All_minorities"))

# inspect
dissimilarity_index_2
```

```{r}
# generate spatial neighbour relationships for Bradford OAs using Queen's contiguity
neighbour_OAs <- poly2nb(bradford_map, queen = TRUE)
# provide summary statistics of the neighbourhood relationships
summary(neighbour_OAs)
```

```{r}
# plot the geographic boundaries of Bradford's Output Areas
plot(bradford_map$geometry)
# overlay neighbourhood connections onto the map
plot(
  neighbour_OAs, # adjacency matrix
  coords = bradford_map$geometry, # plot neighbour connections from polygon centroids
  col ="red", # colour of neighbour lines
  add  = T # add neighbour lines to existing plot
)
```

```{r}
# create neighbour relationships to spatial weights list format
neighbour_OAs <- nb2listw(neighbour_OAs)

# calculate Getis-Ord Gi* statistic for spatial clustering (hot/cold spots) for "White British" group
g_White_British <- localG(bradford_map$p_White_British, neighbour_OAs)

# calculate Getis-Ord Gi* statistic for spatial clustering (hot/cold spots) for "Pakistani" group
g_Pakistani <- localG(bradford_map$p_Pakistani, neighbour_OAs)
```

```{r}
# append calculated Gi* statistic for White British to the spatial data frame
g_map <- cbind(bradford_map, as.matrix(g_White_British)) %>%
    rename(g_White_British = as.matrix.g_White_British.)
```

```{r}
# append calculated Gi* statistic for Pakistani to the spatial data frame
g_map <- cbind(g_map, as.matrix(g_Pakistani)) %>%
         rename(g_Pakistani = as.matrix.g_Pakistani.)
```

```{r}
# create a thematic map to visualise hotspots and coldspots for White British and Pakistani populations
tm_shape(g_map) +  
  tm_fill(c("g_White_British", "g_Pakistani"), # visualise Gi* values for both ethnic groups
         palette = "-RdBu", # diverging palette to highligh hot and cold spots
         style = "fixed", # fixed breaks for significance thresholds
         breaks = c(-10, -2.58, -1.96, 1.96, 2.58, 10), # thresholds for isgnificance (95% and 99%)
         title = c("White British (Gi*)", "Pakistani (Gi*)")) + # titles in the legend
  tm_borders(col = "darkgrey",
             alpha = .25) +
  tm_scale_bar(position = c("LEFT", "BOTTOM")) +
  tm_compass(position = c("RIGHT", "TOP")) +
  tm_layout(main.title = "Hotspots and Coldspots for Ethnic Groups in Bradford OAs",
            main.title.position = c("center", "top"),
            main.title.size = 1.2,
            main.title.fontface = "bold",
            legend.frame = TRUE,
            legend.title.size = 1)
```

Producing Key Output

```{r}
# filter the ethnicity_OAs data to include only rows for Walsall
walsall_OAs <- filter(ethnicity_OAs, Local_Authority == "Walsall")
```

```{r}
# calculate percentage-based ethnicity columns to the Walsall spatial data frame
walsall_summary <- summarise(walsall_OAs,
                             across(White_British:Other_groups, ~ .x/Total_OA_pop*100)
                             ) %>% 
  mutate(rowid = row_number())
```

```{r}
# transform Walsall ethnicity data from wide to long format for easier plotting and analysis
walsall_summary_long <- pivot_longer(walsall_summary, cols = White_British:Other_groups, names_to = "Ethnicity", values_to = "Percent")
```

```{r}
# create a boxplot showing the distribution of ethnicity percentages across Walsall OAs
ggplot(walsall_summary_long, aes(Ethnicity, Percent)) +
  geom_boxplot() + # add boxplots for each ethnic group
  theme_bw() + # apply a clean black-and-white theme
  ggtitle("Distribution of Ethnicity Percentages in Walsall OAs") + # plot title
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 30, vjust = 0.6))
```

```{r}
# calculate summary statistics (mean, median, maximum, minimum) for ethnicity percentages in Walsall OAs
walsall_summary_statistics <- walsall_summary_long %>% 
  group_by(Ethnicity) %>% # group data by Ethnicity category
  summarise(
    mean = mean(Percent), # calculate average percentage
    median = median(Percent), # calculate median percentage
    maximum = max(Percent), # calculate maximum percentage value
    minimum = min(Percent) # calculate minimum percentage
  ) %>% 
  mutate(across(c(mean, median, maximum, minimum), ~ round(.x, 1)))
```

```{r}
# merge shapefile with Walsall's ethnicity data to create a spatial data frame using the OA21CD and OA_code column
# default merging will drop all row which do not appear in both the datasets that are being merged - only the Walsall OAs will be left after merging
walsall_map <- merge(output_area_shapes, walsall_OAs, by.x = "OA21CD", by.y = "OA_code")
```

```{r}
# recalculate percentage-based ethnicity columns to the Walsall spatial data frame
walsall_map <- mutate(walsall_map, across(White_British:Other_groups, ~.x/Total_OA_pop*100,.names = paste("p", "{.col}", sep = "_")))
```

```{r}
# create thematic map visualising percentages of White British and Pakistani groups in Walsall Output Areas
tm_shape(walsall_map) +
  tm_fill(col = c("p_White_British", "p_Indian"), # columns to visualise as percentages
          title = c("% White British", "% Indian"),
          style = "fixed", # use fixed intervals for colour breaks
          breaks = c(0, 10, 20, 50, 100)) + # defined intervals (0-10, 10-20, 20-50, 50-100)
  tm_borders(col = "darkgrey",
             alpha = .25) +
  tm_scale_bar(position = c("center", "bottom")) +
  tm_compass(position = c("right", "top")) +
  tm_layout(main.title = "% of White British and Indian Populations in Walsall OAs",
            main.title.position = c("center", "top"),
            main.title.size = 1.2,
            main.title.fontface = "bold",
            legend.frame = TRUE,
            legend.title.size = 1)
```

```{r}
# calculate the Index of Dissimilarity between Indian and White British populations in Walsall OAs
dissimilarity_index_3 <- id(walsall_OAs, vars = c("Indian", "White_British"))

# inspect
# the dissimilarity value of 0.54 is high but not as much as that between the White British and Pakistani populations in Bradford as we saw earlier in the map
dissimilarity_index_3

# extract the numeric index value
dissimilarity_value <-  as.numeric(dissimilarity_index_3)
```

```{r}
# convert dissimilartiy index into a data frame
dissimilarity_df <- data.frame(
  Ethnic_Groups = "White British & Indian",
  Dissimilarity_Index = round(dissimilarity_value, 2) # round to 2 decimal places
)

# experiment with various packages to convert table output
library(gt)
library(pander)
library(xtable)
library(flextable)
library(stargazer)

# gt
dissimilarity_df %>%
  gt() %>%
  tab_header(title = "Index of Dissimilarity in Walsall")

# pander
pander(dissimilarity_df, caption = "Index of Dissimilarity in Walsall")

# xtable
print(xtable(dissimilarity_df, caption = "Index of Dissimilarity in Walsall"), type = "latex")

# flextable
dissimilarity_df %>%
  flextable() %>%
  set_caption("Index of Dissimilarity in Walsall")

# stargazer
stargazer(as.data.frame(dissimilarity_df), type = "text", summary = FALSE, 
          title = "Index of Dissimilarity in Walsall", rownames = FALSE)
```


```{r}
# generate spatial neighbour relationships for Walsall OAs using Queen's contiguity
neighbour_OAs <- poly2nb(walsall_map, queen = TRUE)
# provide summary statistics of the neighbourhood relationships
summary(neighbour_OAs)
```


```{r}
# plot the geographic boundaries of Bradford's Output Areas
plot(walsall_map$geometry)
# overlay neighbourhood connections onto the map
plot(
  neighbour_OAs, # adjacency matrix
  coords = walsall_map$geometry, # plot neighbour connections from polygon centroids
  col ="red", # colour of neighbour lines
  add  = T # add neighbour lines to existing plot
)
```

```{r}
# create neighbour relationships to spatial weights list format
neighbour_OAs <- nb2listw(neighbour_OAs)

# calculate Getis-Ord Gi* statistic for spatial clustering (hot/cold spots) for "White British" group
g_White_British <- localG(walsall_map$p_White_British, neighbour_OAs)

# calculate Getis-Ord Gi* statistic for spatial clustering (hot/cold spots) for "Indian" group
g_Indian <- localG(walsall_map$p_Indian, neighbour_OAs)
```

```{r}
# append calculated Gi* statistic for White British to the spatial data frame
g_map <- cbind(walsall_map, as.matrix(g_White_British)) %>%
    rename(g_White_British = as.matrix.g_White_British.)
```

```{r}
# append calculated Gi* statistic for Indian to the spatial data frame
g_map <- cbind(g_map, as.matrix(g_Indian)) %>%
         rename(g_Indian = as.matrix.g_Indian.)
```

```{r}
# create a thematic map to visualise hotspots and coldspots for White British and Pakistani populations
tm_shape(g_map) +  
  tm_fill(c("g_White_British", "g_Indian"), # visualise Gi* values for both ethnic groups
         palette = "-RdBu", # diverging palette to highligh hot and cold spots
         style = "fixed", # fixed breaks for significance thresholds
         breaks = c(-10, -2.58, -1.96, 1.96, 2.58, 10), # thresholds for isgnificance (95% and 99%)
         title = c("White British (Gi*)", "Indian (Gi*)")) + # titles in the legend
  tm_borders(col = "darkgrey",
             alpha = .25) +
  tm_scale_bar(position = c("center", "bottom")) +
  tm_compass(position = c("right", "top")) +
  tm_layout(main.title = "Hotspots and Coldspots for Ethnic Groups in Walsall OAs",
            main.title.position = c("center", "top"),
            main.title.size = 1.2,
            main.title.fontface = "bold",
            legend.frame = TRUE,
            legend.title.size = 1)
```

Evidence Brief for Policy Report

For your chosen authority, you then need to answer the following question:

1. To what extent is the area more divided by ethnicity than by occupational social class?

For the occupational class data, there is an additional csv on Moodle containing five variables: OA codes, Local Authority names, the count of people with NSSEC 1-2 occupations (managerial, administrative and professional), the count of people with NSSEC 5-7 occupations (semi/routine and manual jobs) and a final variable listing the total OA population of residents with NSSEC values. You can read more about NSSEC here: https://www.ons.gov.uk/employmentandlabourmarket/peopleinwork/employmentandemployeetypes/bulletins/industryandoccupationenglandandwales/census2021. 

Hints: the approach you take is up to you. You could use statistics and/or maps to answer the question and you will want to focus the ethnic dimension of the brief to keep things manageable (e.g. by picking two relevant groups to compare, by undertaking a 'majority-minority' comparison and so forth). Think carefully about what you think is the most effective way to deliver a succinct answer to the brief and feel free to reach out if you are unsure. 

```{r}
NSSEC_OAs <- read_csv(here("Practical and Evidence Brief resources-20250311", "NSSEC_OA_2021.csv"))
```

```{r}
NSSEC_OAs
```

```{r}
walsall_NSSEC_OAs <- filter(NSSEC_OAs, Local_Authority == "Walsall")
```

```{r}
walsall_NSSEC_summary <- summarise(walsall_NSSEC_OAs, across(NSSEC1_2:NSSEC5_7, ~ .x/OA_total*100)) %>% 
  mutate(rowid = row_number())
```
```{r}
walsall_NSSEC_summary_long <- pivot_longer(walsall_NSSEC_summary, cols = NSSEC1_2:NSSEC5_7, names_to = "NSSEC", values_to = "Percent")
```

```{r}
ggplot(walsall_NSSEC_summary_long, aes(NSSEC, Percent)) + geom_boxplot() + theme_bw()
```

```{r}
bradford_map <- merge(output_area_shapes, bradford_OAs, by.x = "OA21CD", by.y = "OA_code")
```

```{r}
walsall_ <- mutate(bradford_map, across(White_British:Other_groups, ~.x/Total_OA_pop*100,.names = paste("p", "{.col}", sep = "_")))
```

```{r}
tm_shape(bradford_map) + tm_fill(col = c("p_White_British", "p_Pakistani"), title = c("% White British", "% Pakistani"), style = "fixed", breaks = c(0, 10, 20, 50, 100)) + tm_borders(col = "darkgrey", alpha = .25) + tm_scale_bar(position = c("LEFT", "BOTTOM")) + tm_compass(position = c("RIGHT", "TOP"))
```
